---
title: "Tutorial3"
format: html
editor: source
---

# Admin

```{r, message= F}
library(tidyverse)
library(nimble)
library(coda)
```

Assessed Tutorial 3 Tutorial 3 is assessed. The assessed questions (not all are assessed!) should be answered within the corresponding Moodle Quiz, by Wednesday 1 October, 2pm.

The tutorial aims to help you practice producing posterior distributions and summaries using Nimble, and assess the convergence of MCMC algorithms.

Tutorial 3 Moodle Quiz

Answer questions b(i), b(ii), b(iii), b(iv), b(v), b(viii), c(i), c(ii), c(iv), c(vi), using the corresponding Moodle quiz. There are 12 marks overall.

The Quiz will open on Friday 26 September at 9am and close on Wednesday 1 October at 2pm. Note that the quiz only allows 1 final submitted attempt, So, you should only press the Submit all and finish button once. You can revise your answers, but not after you have pressed this button!

Note also that attempts that have not been submitted before the deadline, will be automatically submitted on Wednesday 1 October at 2pm.

# Part A

The data

Consider birth weight data 𝑤 = (𝑤1, ..., 𝑤200), in Kg, from 200 births at Hospital H. The data are given in the R workspace file Birth_weights_Tutorial_3 in the Tutorials folder in Moodle. The data file also contains binary information on the Index of Multiple Deprivation (IMD) of the mother. 0 denotes high deprivation (index from 1 to 5), while 1 denotes low deprivation (index from 6 to 10). Note the counter-intuitive interpretation of the IMD values on the scale from 1 to 10, and subsequently on the binary scale.

```{r, message=FALSE}
load("~/Documents/St. Andrews/Bayes/AppliedBayes/Birth_weights_Tutorial_3.RData")

d <- as.data.frame(birth_weights) |>
  mutate(j = seq(1, length(birth_weights), 1))

d2 <- as.data.frame(IMD_binary)|>
  mutate(j = seq(1, length(IMD_binary), 1))

d <- left_join(d, d2) |>
  mutate(bw = birth_weights,
         imd = IMD_binary,
         imd_plot = ifelse(imd == 1, "Low", "High")) |>
  select(-c(j, birth_weights, IMD_binary))

remove(d2, birth_weights, IMD_binary)
```

i)  Plot the data, both ignoring and taking into account the IMD variable.

```{r}
d |>
  ggplot(aes(x = bw)) + 
  geom_histogram(binwidth = 0.25) +
  theme_bw() +
  labs(x = "Birth Weights", y = "Frequency")
```

```{r}
d |>
  ggplot(aes(x = bw)) +
  geom_histogram(binwidth = 0.25) +
  # coord_flip() +
  facet_grid(~ imd_plot) +
  theme_bw() +
  scale_y_continuous(breaks = NULL) +
  labs(y = "", x = "Birth Weight (Kg)")
```

ii) What is the sample mean of the 200 birth weights?

```{r}
m <- mean(d$bw)

paste0("The mean birth weight in the sample is ", round(m, 3), " kilograms.")
```

iii) What is the sample variance of the 200 observations?

```{r}
v <- var(d$bw)

paste0("The variance of the sample birth weight is ", round(v, 3), " kilograms.")
```

iv) What is the sample mean of the high-deprivation birth weights?

```{r}
m_hd <- mean(d$bw[d$imd == 0])

paste0("The mean birth weight in the high-deprivation sample is ", round(m_hd, 3), " kilograms.")
```

v)  What is the sample mean of the low-deprivation birth weights?

```{r}
m_ld <- mean(d$bw[d$imd == 1])

paste0("The mean birth weight in the low-deprivation sample is ", round(m_ld, 3), " kilograms.")
```

# Part B

```{r}
################################################################
################################################################

# Specify the statistical model
Ncode <- nimbleCode({
  
 # Specify the likelihood:
  for (i in 1:N){
    x[i] ~ dnorm(mu, tau)
  }
 # to generate values for the average waiting time, given the \lambda samples
  # averagewaitingtime<-1/lambda
  
  prior_mu <- 3
  prior_var <- 4
  prior_precision <- 1/prior_var
  
  prior_tau_a <- 0.1
  prior_tau_b <- 0.1
  
 # Prior specification:
  mu ~ dnorm(prior_mu, prior_precision) # Gamma prior with known parameters c and d
  
  tau ~ dgamma(prior_tau_a, prior_tau_b)
})
```

```{r}
# Values for some constants in the model
N_constants <- list(N = 200) # for a prior that is not very informative
# ExpGammaConsts <- list(N = 6, c=100, d=120) # for a very informative prior

# The data values
N_data <- list(x=d$bw)

# For no data at all! Just to see what the prior(s) look like, as a sanity check!
# ExpGammaData <- list(x=c(NA,NA,NA,NA,NA,NA))

# one set of initial values before building the model                 
initials <- list(mu=1, tau = 0.5)  # missing data are random variables 
#                                   and need to be initialised too. 
                                 # Nimble can do this by sampling from the priors. 

# to build the model
Nmod <- nimbleModel(code = Ncode, name = "Nmod", 
        constants = N_constants, data = N_data, inits<-initials)

# To compile the model
CNmod <- compileNimble(Nmod)

# set up the monitored quantities. Default is all of the random quantities
NmodConf <- configureMCMC(Nmod, 
                    monitors = c('mu',"tau"), print = TRUE) 

# build the MCMC algorithm
NmodMCMC <- buildMCMC(NmodConf)
# compile the MCMC chain 
CNmodMCMC <- compileNimble(NmodMCMC, project = Nmod)
```


```{r}
####################################################################################
####### POSTERIOR SAMPLES IN CODA FORMAT TO GET MORE EASILY PLOTS AND DIAGNOSTICS  #
####################################################################################
set.seed(10)
NmodInits <- list(list(mu = 1, tau = 0.5), list(mu = 10, tau = 2))
# note that number of iterations niter contains the number of burn-in samples
posterior <- runMCMC(CNmodMCMC, niter = 10000, thin=1, nburnin=1000,
                   summary = TRUE, WAIC = FALSE, samples = TRUE, nchains=2,
                   samplesAsCodaMCMC = TRUE,inits = NmodInits)

combinedchains <- mcmc.list(posterior$samples$chain1,
                            posterior$samples$chain2)
# plot(combinedchains)
```
i) State the Monte Carlo error for 𝜇. [1 mark]

```{r}
round(batchSE(posterior$samples$chain1,batchSize=100), 5)
```

This output shows that the Monte Carlo error for mu is 0.00052.

ii) State the Effective Sample Size for 𝜎[1 mark]

```{r}
eff_s_s_mu <- effectiveSize(posterior$samples$chain1[ , "mu"])
cat("The effective sample size for mu for Chain 1 is",eff_s_s_mu)
```

```{r}
eff_s_s_tau <- effectiveSize(posterior$samples$chain1[ , "tau"])
cat("The effective sample size for tau for Chain 1 is",eff_s_s_tau)
```

These outputs show the effective sample sizes for both parameters, the effective sample size for tau is 8925.751. 

iii) At least one of the autocorrelation plots provides some cause for concern. Is this statement
true or false? [1 mark]


```{r}
autocorr.plot(posterior$samples$chain1)
```
Neither of these plots show concerning autocorrelation among the samples in chain 1.


```{r}
autocorr.plot(posterior$samples$chain2)

```

Neither of these plots show concerning autocorrelation among the samples in chain 2.

So, the statement that at least one of the autocorrelation plots provides some cause for concern is not correct.

iv) The BGR diagnostics provide cause for concern. Is this statement true or false? [1 mark]


```{r}
gelman.diag(combinedchains)
```

```{r}
gelman.plot(combinedchains)
```

It appears that our chain length is not sufficient for the tau estimation chain to fully converge based on the between and within variances of the chains. The chains for mu do seem to have converged.

So, yes, the BGR diagnostics do provide cause for concern.

v) State the posterior mean for 𝜇 to an accuracy of 3 decimal places. [1 mark]
vi) State the posterior mean for 𝜎to an accuracy of 3 decimal places.

```{r}
b_summary <- as.data.frame(posterior$summary$all.chains)

paste0("The posterior mean birth weight of the entire sample is ", round(b_summary$Mean[1], 3), " kilograms.")

paste0("The posterior variance in birth weight of the entire sample is ", signif(b_summary$Mean[2], 3), "0", " kilograms.")

# had to put the 0 in a character string because the default for the round function is to take a rounded value of 1.790 and only print 1.79.
```

vii) Using R, and for each parameter, plot the prior and posterior distributions together.

```{r}
par(mfrow = c(1, 2))

plot(density(posterior$samples$chain1[ , "mu"]), type = "l", 
  xlab = expression(mu), ylab = "Posterior density", main="Mu Chain 1", xlim = c(2, 4))

mu_forplot = seq(2, 4, length=200)

lines(mu_forplot, dnorm(mu_forplot, 3, 0.25), type='l', col="red")

legend(2, 4, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)


plot(density(posterior$samples$chain2[ , "mu"]), type = "l", 
  xlab = expression(mu), ylab = "Posterior density", main="Mu Chain 2", , xlim = c(2, 4))

mu_forplot = seq(2, 4, length=200)

lines(mu_forplot, dnorm(mu_forplot, 3, 0.25), type='l', col="red")

legend(2, 4, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)
```

```{r}
par(mfrow = c(1, 2))

plot(density(posterior$samples$chain1[ , "tau"]), type = "l", 
  xlab = expression(tau), ylab = "Posterior density", main="Tau Chain 1", xlim = c(0, 3))

tau_forplot = seq(0, 3, length=200)

lines(tau_forplot, dgamma(tau_forplot, 0.1, 0.1), type='l', col="red")

legend(0.4, 1, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)


plot(density(posterior$samples$chain2[ , "tau"]), type = "l", 
  xlab = expression(tau), ylab = "Posterior density", main="Tau Chain 2", xlim = c(0, 3))

tau_forplot = seq(0, 3, length=200)

lines(tau_forplot, dgamma(tau_forplot, 0.1, 0.1), type='l', col="red")

legend(0.4, 1, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)
```

viii) Briefly report your inferences regarding the posterior expected birth weight to someone
with no knowledge of statistics, considering the application at hand. [2 marks]

Based on our analysis of the birth-weight data, we expect that the mean of the distribution of birth weights of babies born in hospital H to be around 3.255 kilograms. (assuming that our sample was an unbiased random sample of the population of babies born at hospital H). We expect the standard deviation of the distribution of birth weights to be 0.75 This relatively high standard deviation indicates that either we do not have a large enough sample size to make precise estimates, or we may be missing an important consideration in this analysis.

# Part C

```{r}
################################################################
################################################################

# Specify the statistical model
Ncode_c <- nimbleCode({
  
 # Specify the likelihood:
  for (i in 1:N){
    # x[i] ~ dnorm((imd[i]-1)*(-mu1) + (diff+mu1)*imd[i], tau)
    x[i] ~ dnorm(mu1 + diff * imd[i], tau)
  }
  
  prior_mu <- 3
  prior_var <- 4
  prior_precision <- 1/prior_var
  
  prior_tau_a <- 0.1
  prior_tau_b <- 0.1
  
 # Prior specification:
  mu1 ~ dnorm(prior_mu, prior_precision) 
  
  diff ~ dnorm(0, prior_precision)
  
  tau ~ dgamma(prior_tau_a, prior_tau_b)
})
```

```{r}
# Values for some constants in the model
N_constants_c <- list(N = 200) 

# The data values
N_data_c <- list(x = d$bw, imd = d$imd)

# one set of initial values before building the model                 
initials_c <- list(mu1 = 1, diff = 0.5, tau = 0.5)  

# missing data are random variables and need to be initialised too. 
 # Nimble can do this by sampling from the priors. 

# to build the model
Nmod_c <- nimbleModel(code = Ncode_c, name = "Nmod_c", 
        constants = N_constants_c, data = N_data_c, inits<-initials_c)

# To compile the model
CNmod_c <- compileNimble(Nmod_c)

# set up the monitored quantities. Default is all of the random quantities
NmodConf_c <- configureMCMC(Nmod_c, 
                    monitors = c('mu1', 'diff', "tau"), print = TRUE) 

# build the MCMC algorithm
NmodMCMC_c <- buildMCMC(NmodConf_c)
# compile the MCMC chain 
CNmodMCMC_c <- compileNimble(NmodMCMC_c, project = Nmod_c)
```

```{r}
####################################################################################
####### POSTERIOR SAMPLES IN CODA FORMAT TO GET MORE EASILY PLOTS AND DIAGNOSTICS  #
####################################################################################
set.seed(10)
NmodInits_c <- list(list(mu1 = 3, diff = 0.3, tau = 1.8), list(mu1 = 3, diff = 0.3, tau = 1.8))
# note that number of iterations niter contains the number of burn-in samples
posterior_c <- runMCMC(CNmodMCMC_c, niter = 10000, thin=1, nburnin=1000, 
                   summary = TRUE, WAIC = FALSE, samples = TRUE, nchains=2, 
                   samplesAsCodaMCMC = TRUE,inits = NmodInits_c) 

combinedchains <- mcmc.list(posterior_c$samples$chain1, 
                            posterior_c$samples$chain2)

png("combinedchains.png", width = 1200, height = 800)
plot(combinedchains)
dev.off()
```

i) At least one of the autocorrelation plots provides some cause for concern. Is this statement
True or False? [1 mark]

```{r}
autocorr.plot(posterior_c$samples$chain1)
```

For both the mu and diff chains, these autocorrelation plots show that there are issues with autocorreltaion of observations in the mu and diff estimation chains when the lag is less than 5. 

```{r}
autocorr.plot(posterior_c$samples$chain2)
```

For both the mu and diff chains, these autocorrelation plots show that there are issues with autocorreltaion of observations in the mu and diff estimation chains when the lag is less than 5.

So, yes, I would say that at least one of the autocorrelation graphs shows cause for some concern.

ii) The BGR diagnostics provide cause for concern. Is this statement True or False? [1
mark]

```{r}
gelman.diag(combinedchains)
```

```{r}
gelman.plot(combinedchains)
```

There is nothing concerning in these plots, all chains seem to converge well.

iii) State the posterior expectation for the mean birth weight of the high-deprivation group to
an accuracy of 2 decimal places.

```{r}
summary_c <- as.data.frame(posterior_c$summary$all.chains)

paste0("The posterior mean birth weight of the high-deprevation group is ", 
       round(summary_c$Mean[2], 2), "0 kilograms.")

# had to put the 0 in a character string because the default for the round function is to take a rounded value of 3.10 and only print 3.1
```

iv) State the posterior expectation for the mean birth weight of the low-deprivation group to
an accuracy of 2 decimal places. [1 mark]

```{r}
paste0("The posterior variance in birth weight of the low-deprevation group is ", 
       round(summary_c$Mean[1] + summary_c$Mean[2], 2), " kilograms.")
```

v) Using R, and for each parameter, plot the prior and posterior distributions together.

```{r}
  prior_mu <- 3
  prior_var <- 4
  prior_precision <- 1/prior_var
  
  prior_tau_a <- 0.1
  prior_tau_b <- 0.1
  
 # Prior specification:
  mu1 ~ dnorm(prior_mu, prior_precision) 
  
  diff ~ dnorm(prior_mu, prior_precision)
  
  tau ~ dgamma(prior_tau_a, prior_tau_b)
```


```{r}
par(mfrow = c(1, 2))

plot(density(posterior_c$samples$chain1[ , "mu1"]), type = "l", 
  xlab = expression(mu), ylab = "Posterior density", main="Mu1 Chain 1", xlim = c(2, 4))

mu1_forplot = seq(2, 4, length=200)

lines(mu1_forplot, dnorm(mu1_forplot, prior_mu, prior_precision), type='l', col="red")

legend(2, 2, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)


plot(density(posterior_c$samples$chain2[ , "mu1"]), type = "l", 
  xlab = expression(mu), ylab = "Posterior density", main="Mu1 Chain 2", xlim = c(2, 4))

mu1_forplot = seq(2, 4, length=200)

lines(mu1_forplot, dnorm(mu1_forplot, prior_mu, prior_precision), type='l', col="red")

legend(2, 2, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)
```

```{r}
par(mfrow = c(1, 2))

plot(density(posterior_c$samples$chain1[ , "diff"]), type = "l", 
  xlab = "Diff", ylab = "Posterior density", main="Diff Chain 1", xlim = c(-1, 4))

diff_forplot = seq(-1, 4, length=200)

lines(diff_forplot, dnorm(diff_forplot, prior_mu, prior_precision), type='l', col="red")

legend(2, 3, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)


plot(density(posterior_c$samples$chain2[ , "diff"]), type = "l", 
  xlab = "Diff", ylab = "Posterior density", main="Diff Chain 2", xlim = c(-1, 4))

diff_forplot = seq(-1, 4, length=200)

lines(diff_forplot, dnorm(diff_forplot, prior_mu, prior_precision), type='l', col="red")

legend(2, 3, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)
```

```{r}
par(mfrow = c(1, 2))

plot(density(posterior_c$samples$chain1[ , "tau"]), type = "l", 
  xlab = expression(tau), ylab = "Posterior density", main="Tau Chain 1", xlim = c(0, 3))

tau_forplot = seq(0, 3, length=200)

lines(tau_forplot, dgamma(tau_forplot, 0.1, 0.1), type='l', col="red")

legend(0.3, 0.8, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)


plot(density(posterior_c$samples$chain2[ , "tau"]), type = "l", 
  xlab = expression(tau), ylab = "Posterior density", main="Tau Chain 2", xlim = c(0, 3))

tau_forplot = seq(0, 3, length=200)

lines(tau_forplot, dgamma(tau_forplot, 0.1, 0.1), type='l', col="red")

legend(0.3, 0.8, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)
```

vi) Considering the statistical output, is it preferable to fit a model with only one overall mean
for the 200 observations, or a model with a different mean for the two deprivation groups?
Briefly justify your answer. [2 marks]

```{r}
eff_s_s_mu1 <- effectiveSize(posterior_c$samples$chain1[ , "mu1"])
cat("The effective sample size for mu for chain1 is",eff_s_s_mu1)
```

```{r}
eff_s_s_diff <- effectiveSize(posterior_c$samples$chain1[ , "diff"])
cat("The effective sample size for the difference in means for chain1 is",eff_s_s_diff)
```

```{r}
eff_s_s_tau <- effectiveSize(posterior_c$samples$chain1[ , "tau"])
cat("The effective sample size for tau for chain1 is",eff_s_s_tau)
```

```{r}
batchSE(posterior_c$samples$chain1,batchSize=100) # For MC errors (only one chain is used.)
```

```{r}
summary_c
```

```{r}
b_summary
```

Based on the required and additional analysis of these data, I think that the approach with two different groups is preferable. While the estimated variance of the sample is slightly higher (lower tau value), and there is no real difference in the values of the standard deviations of the estimates of the parameters in each model, the fact that during exploratory data analysis, we found a clear difference in the means of the high and low deprevation groups, I think that the approach estimating two group means is a more appropriate approach to this problem.

