---
title: "Tutorial5"
format: html
editor: 
  markdown: 
    wrap: 80
---

# Packages and Data

```{r, message= F}
library(tidyverse)
library(nimble)
library(coda)
library(igraph)
library(R6)
library(knitr)
```

Consider birth weight data ğ‘¤ = (ğ‘¤1, ..., ğ‘¤200), in Kg, from 200 births at
Hospital H. The data are given in the R workspace file Birth_weights_Tutorial_3
in the Tutorials folder in Moodle. The data file also contains binary
information on the Index of Multiple Deprivation (IMD) of the mother. 0 denotes
high deprivation (index from 1 to 5), while 1 denotes low deprivation (index
from 6 to 10). Note the counter-intuitive interpretation of the IMD values on
the scale from 1 to 10, and subsequently on the binary scale.

```{r, message=FALSE}
load("Birth_weights_Tutorial_5.RData")

d <- as.data.frame(cbind(age, birth_weights, IMD_binary, pm2_5))

d <- d |>
  rename(
    bw = birth_weights,
    imd = IMD_binary) |>
  mutate(
    cage = age - mean(age),
    cpm2_5 = pm2_5 - mean(pm2_5),
    imd_plot = ifelse(imd == 1, "Low", "High"))

# remove(age, birth_weights, cage, cpm2_5, imd, IMD_binary)
```

```{r}
# sanity check frequentist linear model

# lm <- lm(bw ~ imd + cage + cpm2_5, data = d)
# 
# summary(lm)
```

Fit a linear model with the birth weights ğ‘¤ğ‘– , ğ‘– = 1, ..., 200 as the outcome
that follows a Normal distribution so that (given ğœ‡ğ‘– and ğœ), (ğ‘¤ğ‘– \|ğœ‡ğ‘– , ğœ âˆ¼ ğ‘(ğœ‡ğ‘–
, ğœ2 )), with the weights being independent. Center the continuous covariates
age and pm2_5, creating two new covariates, say, **cage** and **cpm2_5**
Overall, the covariates should be the binary IMD variable, as well as the
centered **cage** and **cpm2_5**.

Parametrise the mean ğœ‡ğ‘– for birth weight ğ‘¤ğ‘– so that,

ğœ‡ğ‘– = ğ›½0 + ğ›½1 âˆ— ğ¼ğ‘€ğ·ğ‘– + ğ›½2 âˆ— ğ‘ğ‘ğ‘”ğ‘’ + ğ›½3 âˆ— ğ‘ğ‘ğ‘š2_5.

Place a ğ¿ğ‘œğ‘” âˆ’ ğ‘ğ‘œğ‘Ÿğ‘šğ‘ğ‘™(1, 0.72 ) prior on ğ›½0 . Place suitably non-informative
priors on ğ›½1 , ğ›½2 and ğ›½3 . For the precision ğœ = 1/ğœ2 , we will assume a Gamma
prior distribution ğº(0.1, 0.1).

Use Nimble to derive the posterior distribution ğœ‹(ğ›½0 , ğ›½1 , ğ›½2 , ğ›½3 , ğœ2 \|ğ‘¤).
Run 2 chains. Consider a burn-in of 100000 iterations, with 400000 further
iterations for each chain. Thin by 10.

Obtain plots of the posterior distributions and diagnostic plots. Check the MCMC
convergence, not forgetting to check the chains length in relation to the
autocorrelations and MC error.

Answer the following questions, to an accuracy of 3 decimal places where
appropriate. (Note that you should also be checking model assumptions and model
fit before reporting inferences, but this will be one of the topics in the next
tutorial sheet.)

```{r}
mean_age <- mean(d$age)
pred_age <- 29-mean_age

mean_pm <- mean(d$pm2_5)
pred_pm <- 16-mean_pm

prec <- 1/0.49
```

```{r}
#covariates and response

y = d$bw
imd = d$imd
cage = d$cage
cpm2_5 = d$cpm2_5

# Specify the statistical model

n = nrow(d)

mod <- nimbleCode({

# Specify the likelihood:
for(i in 1:n){
  
  mu[i] <- beta0 + beta1 * imd[i] + beta2 * cage[i] + beta3 * cpm2_5[i]
  
  y[i] ~ dnorm(mu[i],tau)
  
}

# Prior for beta

beta0 ~ dlnorm(1, 2.040816)

beta1 ~ dnorm(mu0,tau0)
beta2 ~ dnorm(mu0,tau0)
beta3 ~ dnorm(mu0,tau0)

tau0 <- 1/sigma02

obs <- beta0 + beta1 * (1) + beta2 * (-1.18) + beta3 * (5.117716)

# Prior for the precision
tau ~ dgamma(a, b)

# Compute the variance
sigma2 <- 1/tau
})
```

```{r}
# hyperparameters for the betas and tau
# would find n in R environment. Including here to suppress a warning
Consts <- list(mu0=0, sigma02=1000, a=0.1, b=0.1, n=200)

# list with data
Data <- list(y=y,imd=imd,cage=cage,cpm2_5=cpm2_5)

# Initial values
Inits <- list(beta0 = 1, beta1 = 0, beta2 = 0, beta3 = 0, tau = 1, obs = 2) 
# Nimble will generate the rest 

# Build and Compile the model
# to build the model
mod_re <- nimbleModel(code = mod, name = "mod", constants = Consts,
                    data = Data, inits = Inits)

# To compile the model
Cmod_re <- compileNimble(mod_re)
```

```{r}
mod_Conf <- 
  configureMCMC(mod_re,
                monitors = c("beta0", "beta1", "beta2", "beta3", "sigma2", "obs"),
                enableWAIC = TRUE, print = TRUE) 
```

```{r}
# build the MCMC algorithm
mod_reMCMC <- buildMCMC(mod_Conf)
# compile the MCMC chain 
Cmod_reMCMC <- compileNimble(mod_reMCMC, project = mod_re)
```

```{r}
####################################################################################
####### POSTERIOR SAMPLES IN CODA FORMAT TO GET MORE EASILY PLOTS AND DIAGNOSTICS  #
####################################################################################
set.seed(1)

mod_reInits <- list(list(beta0 = 3, beta1 = 0, beta2 = 0, beta3 = 0, tau = 2, obs = 2), 
                     list(beta0 = 3, beta1 = 0, beta2 = 0, beta3 = 0, tau = 2, obs = 2))

posterior <- runMCMC(Cmod_reMCMC, niter = 500000, thin=10, nburnin=100000, 
                     summary = TRUE, WAIC = TRUE, samples = TRUE, nchains=2, 
                     samplesAsCodaMCMC=TRUE, inits = mod_reInits) 

combinedchains <- mcmc.list(posterior$samples$chain1, posterior$samples$chain2)
```

```{r}
summary(combinedchains)
```

\

```{r}
post_1 <- as.data.frame(posterior$samples$chain1)
post_2 <- as.data.frame(posterior$samples$chain2)

plot(combinedchains) # too many plots sometimes
```

```{r}
#plot(combinedchains[,c('beta1','beta2','sigma2')]) 
#plot(combinedchains[,'beta[1]']) # when defining a vector of parameters

autocorr.plot(posterior$samples$chain1)
autocorr.plot(posterior$samples$chain1[,"beta0"])

autocorr.plot(posterior$samples$chain2)
```

```{r}
gelman.diag(combinedchains, multivariate = F)
gelman.plot(combinedchains)
```

```{r}
ESS <- t(effectiveSize(posterior$samples$chain2))
ESS <- as.data.frame(ESS)
kable(ESS)
# cat("The Effective Sample Size is ", ESS)

effectiveSize(posterior$samples$chain1)
effectiveSize(posterior$samples$chain2)
```

```{r}
beta0=c(posterior$samples$chain1[,'beta0'],posterior$samples$chain2[,'beta0'])
beta1=c(posterior$samples$chain1[,'beta1'],posterior$samples$chain2[,'beta1'])
beta2=c(posterior$samples$chain1[,'beta2'],posterior$samples$chain2[,'beta2'])
beta3=c(posterior$samples$chain1[,'beta3'],posterior$samples$chain2[,'beta3'])
sigma2=c(posterior$samples$chain1[,'sigma2'],posterior$samples$chain2[,'sigma2'])
sigma=sqrt(sigma2)

niterf=length(beta1)

x=cbind(rep(1,n),d$imd,d$cage,d$cpm2_5)
H=x%*%solve((t(x)%*%x))%*%t(x)

#fitted values
fittedvalues=matrix(0,nrow=n,ncol=niterf)
for(l in 1:niterf){
fittedvalues[,l]=beta0[l]*x[,1]+beta1[l]*x[,2]+beta2[l]*x[,3]+beta3[l]*x[,4]
}

#studentised residuals
studentisedred=matrix(0,nrow=n,ncol=niterf)
for(l in 1:niterf){
for(i in 1:n){
studentisedred[i,l]=(y[i]-fittedvalues[i,l])/(sigma[l]*sqrt((1-diag(H)[i]))) 
}  
}  

#posterior mean of studentised residuals
studentisedredm=numeric(n)
for(i in 1:n){
studentisedredm[i]=mean(studentisedred[i,])  
}

#QQ-plot
qqnorm(studentisedredm,xlim=c(-3,3),ylim=c(-3,3),lwd=2)
qqline(studentisedredm,col=2,lwd=2)


#checking independence of error terms
plot(seq_along(studentisedredm),studentisedredm,xlab="Index",
     ylab="Bayesian studentised residual",ylim=c(-3,3))


#posterior mean fitted values
fittedvaluesm=numeric(n)
for(i in 1:n){
fittedvaluesm[i]=mean(fittedvalues[i,])  
}

plot(fittedvaluesm,studentisedredm,xlab="Fitted value (posterior mean)",
     ylab="Bayesian Studentised residual (posterior mean)")
```

```{r}
# question 2

# need post-pred dist for y[i]= beta0

# beta0 <- as.matrix(posterior$samples$chain1[,1])
# beta1 <- as.matrix(posterior$samples$chain1[,2])
# beta2 <- as.matrix(posterior$samples$chain1[,3])
# beta3 <- as.matrix(posterior$samples$chain1[,4])
# 
# vals <- rep(0,length(beta0))
# 
# ex_cage <- 29 - mean(d$age)
# ex_pm2_5 <- 16 - mean(d$pm2_5)
# 
# for(l in 1:length(beta0)){
# vals[l]= beta0[l] + beta1[l] * 1 + beta2[l] * ex_cage + beta3[l] * ex_pm2_5
# }
# 
# m <- mean(vals)
# m
# 
# quantile(vals, probs = c(0.025, 0.975))

# crit <- 1.96*(sqrt(var(vals))/length(beta0))
# 
# lower <- m - crit
# lower
# 
# upper <- m + crit
# upper
# 
# hist(vals)
```

1.  Posterior inferences can be trusted, considering convergence and mixing. Is
    this statement true or false? \[1 mark\]

    1.  True

2.  Considering births at hospital H in general, state the estimated mean birth
    weight for parents classified within then high-deprivation band, of average
    age, with an exposure to PM2.5 that is also average. \[1 mark\]

    1.  3.218

3.  Provide the left limit of the 95% credible interval for the estimate in (b).
    \[1 mark\]

    1.  3.135

4.  Provide the right limit of the 95% credible interval for the estimate in
    (b). \[1 mark\]

    1.  3.302

5.  State the predicted weight for a specific birth from a mother that is
    classified within the low-deprivation band, is 29 years old, and their
    exposure to PM2.5 is 16 ğœ‡ğ‘”/ğ‘š3. \[1mark\]

    1.  2.447

6.  Provide the left limit of the 95% credible interval for the prediction in
    (e). \[1 mark\]

    1.  2.319

7.  Provide the right limit of the 95% credible interval for the prediction in
    (e). \[1 mark\]

    1.  2.575

8.  Describe to someone with no statistical knowledge the effect of the model
    covariates, considering the interpretation of the estimates for ğ›½0, ğ›½1, ğ›½2,
    ğ›½3. \[3 marks\]

    1.  This model is a multiple linear regression, which is a statistical model
        designed to find the associations between several predictor variables
        and some outcome. In this case, we are trying to find the associations
        between birth weight and several predictors about the mother of the
        baby. These predictors are whether the mother is in the high or low
        group for IMD, a metric about financial deprivation of the mother, the
        age of the mother, and her exposure to pm2.5 particles. These three
        variables join an intercept, representing our expected birth weight if
        the mother was in the high deprivation group and had average age and
        pm2.5 exposure relative to the rest of the sample, in our model,
        allowing us to consider the associations between each of those variables
        and birth weight, while holding the other variables constant.

    2.  The point estimate for the value of that intercept, $\beta_0 = 3.218$,
        indicates that a baby born to a mother in the high deprivation group and
        had average age and pm2.5 exposure relative to the rest of the sample,
        is expected to weigh 3.218 kilograms on average.

    3.  The point estimate for $\beta_1 = 0.075$ indicates that babies born to
        mothers in the high deprevation group are expected to be, on average,
        0.075 kilograms heavier than babies born to mothers in the
        low-deprivation group, holding age and pm2.5 exposure constant.

    4.  The point estimate for $\beta_2 = 0.007$ indicates that for each
        additional year older the mother is, their babies are expected to be, on
        average, 0.007 kilograms heavier, holding IMD and pm2.5 exposure
        constant.

    5.  The point estimate for $\beta_1 = -0.164$ indicates that for each
        additional microgram per cubic meter of pm2.5 particulate in the
        mother's home post code, their babies are expected to be 0.164 kilograms
        lighter, holding age and IMD constant.

```{r}
mean(beta3)
quantile(beta2, probs = c(0.025, 0.975))
```

1.  Comment on how important each one of the covariates seems to be for the
    modelling of birth weights. \[2 marks\]

    1.  Beta1 is the parameter representing the association of the binary
        deprivation group variable with birth weight independently of age and
        exposure to pm2.5. The mean of our posterior samples is 0.075 and the
        95% credible interval of this estimate from chain 1: (-0.044, 0.194)
        represents our uncertainty about this parameter value. Since that
        interval contains 0, meaning that we can not be sure if the parameter
        value is positive or negative, we would not consider this to be one of
        our more important parameters.

    2.  Beta2 is the parameter representing the association between age and
        birth weight independently of IMD and exposure to pm2.5. The mean of our
        posterior samples is 0.007 and the 95% credible interval of this
        estimate from chain 1: (-0.010, 0.024) represents our uncertainty about
        this parameter value, since that interval contains 0, meaning that we
        can not be sure if the parameter value is positive or negative, we would
        not consider this to be one of our more important parameters.

    3.  Beta 3 is the parameter representing the association between exposure to
        pm2.5 and birth weight independently of age and IMD. The mean of our
        posterior samples is -0.164 and the 95% credible interval of this
        estimate from chain 1: (-0.180, -0.148) represents our uncertainty about
        this parameter value, since this interval does not include 0, we are
        inclined to consider this as a more important predictor of birth weight
        than $\beta_1$ or $\beta_2$.
