---
title: "Tutorial3"
format: html
editor: source
---

# Admin

```{r}
library(tidyverse)
library(nimble)
library(coda)
```

Assessed Tutorial 3 Tutorial 3 is assessed. The assessed questions (not all are assessed!) should be answered within the corresponding Moodle Quiz, by Wednesday 1 October, 2pm.

The tutorial aims to help you practice producing posterior distributions and summaries using Nimble, and assess the convergence of MCMC algorithms.

Tutorial 3 Moodle Quiz

Answer questions b(i), b(ii), b(iii), b(iv), b(v), b(viii), c(i), c(ii), c(iv), c(vi), using the corresponding Moodle quiz. There are 12 marks overall.

The Quiz will open on Friday 26 September at 9am and close on Wednesday 1 October at 2pm. Note that the quiz only allows 1 final submitted attempt, So, you should only press the Submit all and finish button once. You can revise your answers, but not after you have pressed this button!

Note also that attempts that have not been submitted before the deadline, will be automatically submitted on Wednesday 1 October at 2pm.

# EDA

The data

Consider birth weight data 𝑤 = (𝑤1, ..., 𝑤200), in Kg, from 200 births at Hospital H. The data are given in the R workspace file Birth_weights_Tutorial_3 in the Tutorials folder in Moodle. The data file also contains binary information on the Index of Multiple Deprivation (IMD) of the mother. 0 denotes high deprivation (index from 1 to 5), while 1 denotes low deprivation (index from 6 to 10). Note the counter-intuitive interpretation of the IMD values on the scale from 1 to 10, and subsequently on the binary scale.

```{r, message=FALSE}
load("~/Documents/St. Andrews/Bayes/AppliedBayes/Birth_weights_Tutorial_3.RData")

d <- as.data.frame(birth_weights) |>
  mutate(j = seq(1, length(birth_weights), 1))

d2 <- as.data.frame(IMD_binary)|>
  mutate(j = seq(1, length(IMD_binary), 1))

d <- left_join(d, d2) |>
  mutate(bw = birth_weights,
         imd = IMD_binary,
         imd_plot = ifelse(imd == 1, "Low", "High")) |>
  select(-c(j, birth_weights, IMD_binary))

remove(d2, birth_weights, IMD_binary)
```

i)  Plot the data, both ignoring and taking into account the IMD variable.

```{r}
d |>
  ggplot(aes(x = bw)) + 
  geom_histogram(binwidth = 0.25) +
  theme_bw() +
  labs(x = "Birth Weights", y = "Frequency")
```

```{r}
d |>
  ggplot(aes(x = bw)) +
  geom_histogram(binwidth = 0.25) +
  # coord_flip() +
  facet_grid(~ imd_plot) +
  theme_bw() +
  scale_y_continuous(breaks = NULL) +
  labs(y = "", x = "Birth Weight (Kg)")
```

ii) What is the sample mean of the 200 birth weights?

```{r}
m <- mean(d$bw)

paste0("The mean birth weight in the sample is ", round(m, 3), " kilograms.")
```

iii) What is the sample variance of the 200 observations?

```{r}
v <- var(d$bw)

paste0("The variance of the sample birth weight is ", round(v, 3), " kilograms.")
```

iv) What is the sample mean of the high-deprivation birth weights?

```{r}
m_hd <- mean(d$bw[d$imd == 0])

paste0("The mean birth weight in the high-deprivation sample is ", round(m_hd, 3), " kilograms.")
```

v)  What is the sample mean of the low-deprivation birth weights?

```{r}
m_ld <- mean(d$bw[d$imd == 1])

paste0("The mean birth weight in the low-deprivation sample is ", round(m_ld, 3), " kilograms.")
```

# Analysis

```{r}
# Revisit the first example in Section 1.2.2 of your notes. 
# X[i] is the waiting time until speaking to a human adviser,
# i, i=1,...,n. 
# Exponential likelihood given \lambda. X[i]~Exp(\lambda)
# 1/lambda is the average waiting time
# Assume n=6, with \xbar=0.9
# Gamma prior on \lambda

################################################################
################################################################

# Specify the statistical model
Ncode <- nimbleCode({
  
 # Specify the likelihood:
  for (i in 1:N){
    x[i] ~ dnorm(mu, tau)
  }
 # to generate values for the average waiting time, given the \lambda samples
  # averagewaitingtime<-1/lambda
  
  prior_mu <- 3
  prior_var <- 4
  prior_precision <- 1/prior_var
  
  prior_tau_a <- 0.1
  prior_tau_b <- 0.1
  
 # Prior specification:
  mu ~ dnorm(prior_mu, prior_precision) # Gamma prior with known parameters c and d
  
  tau ~ dgamma(prior_tau_a, prior_tau_b)
})
```

```{r}
# Values for some constants in the model
N_constants <- list(N = 200) # for a prior that is not very informative
# ExpGammaConsts <- list(N = 6, c=100, d=120) # for a very informative prior

# The data values
N_data <- list(x=d$bw)

# For no data at all! Just to see what the prior(s) look like, as a sanity check!
# ExpGammaData <- list(x=c(NA,NA,NA,NA,NA,NA))

# one set of initial values before building the model                 
initials <- list(mu=1, tau = 0.5)  # missing data are random variables 
#                                   and need to be initialised too. 
                                 # Nimble can do this by sampling from the priors. 

# to build the model
Nmod <- nimbleModel(code = Ncode, name = "Nmod", 
        constants = N_constants, data = N_data, inits<-initials)

# To compile the model
CNmod <- compileNimble(Nmod)

# set up the monitored quantities. Default is all of the random quantities
NmodConf <- configureMCMC(Nmod, 
                    monitors = c('mu',"tau"), print = TRUE) 

# build the MCMC algorithm
NmodMCMC <- buildMCMC(NmodConf)
# compile the MCMC chain 
CNmodMCMC <- compileNimble(NmodMCMC, project = Nmod)
```

```{r}
########################################################################
####### POSTERIOR SAMPLES AS R MATRICES   ##############################
########################################################################
set.seed(10)
NmodInits <- list(list(mu = 1, tau = 0.5), list(mu = 10, tau = 2))
# note that number of iterations niter contains the number of burn-in samples
posterior <- runMCMC(CNmodMCMC, niter = 10000, thin=1, nburnin=1000, 
                   summary = TRUE, WAIC = FALSE, samples = TRUE, nchains=2, 
                   inits = NmodInits) 
```

```{r}
par(mfrow = c(1, 1))
plot(posterior$samples$chain1[ , "mu"], type = "l", xlab = "iteration",
     ylab = expression(mu))
lines(posterior$samples$chain2[ , "mu"], type = "l", xlab = "iteration", col=2, 
     ylab = expression(mu))
```

```{r}
par(mfrow = c(1, 1))
plot(posterior$samples$chain1[ , "tau"], type = "l", xlab = "iteration",
     ylab = expression(tau))
lines(posterior$samples$chain2[ , "tau"], type = "l", xlab = "iteration", col=2, 
     ylab = expression(tau))
```


```{r}
par(mfrow = c(2, 2))
# plot autocorrelation of lambda sample - chain 1
acf(posterior$samples$chain1[, "mu"]) 

acf(posterior$samples$chain2[, "mu"]) 

acf(posterior$samples$chain1[, "tau"]) 

acf(posterior$samples$chain2[, "tau"]) 
```

```{r}
par(mfrow = c(1, 2))

plot(density(posterior$samples$chain1[ , "mu"]), type = "l", 
  xlab = expression(mu), ylab = "Posterior density", main="Mu")

mu_forplot = seq(0, 4, length=200)

lines(mu_forplot, dnorm(mu_forplot, 3, 0.25), type='l', col="red")

legend(2, 0.8, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)


plot(density(posterior$samples$chain2[ , "mu"]), type = "l", 
  xlab = expression(mu), ylab = "Posterior density", main="Mu")

mu_forplot = seq(0, 4, length=200)

lines(mu_forplot, dnorm(mu_forplot, 3, 0.25), type='l', col="red")

legend(2, 0.8, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)
```

```{r}
par(mfrow = c(1, 2))

plot(density(posterior$samples$chain1[ , "tau"]), type = "l", 
  xlab = expression(tau), ylab = "Posterior density", main="Tau")

tau_forplot = seq(0, 4, length=200)

lines(tau_forplot, dgamma(tau_forplot, 0.1, 0.1), type='l', col="red")

legend(2, 0.8, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)


plot(density(posterior$samples$chain2[ , "tau"]), type = "l", 
  xlab = expression(tau), ylab = "Posterior density", main="Tau")

tau_forplot = seq(0, 4, length=200)

lines(tau_forplot, dgamma(tau_forplot, 0.1, 0.1), type='l', col="red")

legend(2, 0.8, legend=c("Prior", "Posterior"),col=c("red", "black"),
        lty = c(1,1), cex=0.5)
```

```{r}
posterior$summary$all.chains
```

```{r}
####################################################################################
####### POSTERIOR SAMPLES IN CODA FORMAT TO GET MORE EASILY PLOTS AND DIAGNOSTICS  #
####################################################################################
set.seed(10)
NmodInits <- list(list(mu = 1, tau = 0.5), list(mu = 10, tau = 2))
# note that number of iterations niter contains the number of burn-in samples
posterior <- runMCMC(CNmodMCMC, niter = 10000, thin=1, nburnin=1000, 
                   summary = TRUE, WAIC = FALSE, samples = TRUE, nchains=2, 
                   samplesAsCodaMCMC = TRUE,inits = NmodInits) 

combinedchains <- mcmc.list(posterior$samples$chain1, 
                            posterior$samples$chain2)
plot(combinedchains)

autocorr.plot(posterior$samples$chain1)

autocorr.plot(posterior$samples$chain2)
```

```{r}
gelman.diag(combinedchains)
```

```{r}
gelman.plot(combinedchains)
```

```{r}
posterior$summary$all.chains
```

```{r}
eff_s_s_mu <- effectiveSize(posterior$samples$chain1[ , "mu"])
cat("The effective sample size for mu for chain1 is",eff_s_s_mu)
```
```{r}
eff_s_s_tau <- effectiveSize(posterior$samples$chain1[ , "tau"])
cat("The effective sample size for tau for chain1 is",eff_s_s_tau)
```

```{r}
batchSE(posterior$samples$chain1,batchSize=100) # For MC errors (only one chain is used.)
```




# Part C

From Prof: set up a model here with two mean parameters, one which is the mean of the high deprevation group, the other is the difference between the two means

```{r}
# Revisit the first example in Section 1.2.2 of your notes. 
# X[i] is the waiting time until speaking to a human adviser,
# i, i=1,...,n. 
# Exponential likelihood given \lambda. X[i]~Exp(\lambda)
# 1/lambda is the average waiting time
# Assume n=6, with \xbar=0.9
# Gamma prior on \lambda

################################################################
################################################################

# Specify the statistical model
Ncode <- nimbleCode({
  
 # Specify the likelihood:
  for (i in 1:N){
    # x[i] ~ dnorm((imd[i]-1)*(-mu1) + (diff+mu1)*imd[i], tau)
    x[i] ~ dnorm(mu1 + diff * imd[i], tau)
  }
  

  #### 
  
  #Does this model then treat mu1 as a constant when imd[i] = 1? how does it estimate the diff parameter if it comes first in the dataset/early in the dataset. Under the hood question
  
  ####
  
  prior_mu <- 3
  prior_var <- 4
  prior_precision <- 1/prior_var
  
  prior_tau_a <- 0.1
  prior_tau_b <- 0.1
  
 # Prior specification:
  mu1 ~ dnorm(prior_mu, prior_precision) # Gamma prior with known parameters c and d
  
  diff ~ dnorm(prior_mu, prior_precision)
  
  tau ~ dgamma(prior_tau_a, prior_tau_b)
})
```

```{r}
# Values for some constants in the model
N_constants <- list(N = 200) # for a prior that is not very informative
# ExpGammaConsts <- list(N = 6, c=100, d=120) # for a very informative prior

# The data values
N_data <- list(x = d$bw, imd = d$imd)

# For no data at all! Just to see what the prior(s) look like, as a sanity check!
# ExpGammaData <- list(x=c(NA,NA,NA,NA,NA,NA))

# one set of initial values before building the model                 
initials <- list(mu1 = 1, diff = 0.5, tau = 0.5)  # missing data are random variables 
#                                   and need to be initialised too. 
                                 # Nimble can do this by sampling from the priors. 

# to build the model
Nmod <- nimbleModel(code = Ncode, name = "Nmod", 
        constants = N_constants, data = N_data, inits<-initials)

# To compile the model
CNmod <- compileNimble(Nmod)

# set up the monitored quantities. Default is all of the random quantities
NmodConf <- configureMCMC(Nmod, 
                    monitors = c('mu1', 'diff', "tau"), print = TRUE) 

# build the MCMC algorithm
NmodMCMC <- buildMCMC(NmodConf)
# compile the MCMC chain 
CNmodMCMC <- compileNimble(NmodMCMC, project = Nmod)
```
```{r}
####################################################################################
####### POSTERIOR SAMPLES IN CODA FORMAT TO GET MORE EASILY PLOTS AND DIAGNOSTICS  #
####################################################################################
set.seed(10)
NmodInits <- list(list(mu1 = 3, diff = 0.3, tau = 1.8), list(mu1 = 3, diff = 0.3, tau = 1.8))
# note that number of iterations niter contains the number of burn-in samples
posterior <- runMCMC(CNmodMCMC, niter = 10000, thin=1, nburnin=1000, 
                   summary = TRUE, WAIC = FALSE, samples = TRUE, nchains=2, 
                   samplesAsCodaMCMC = TRUE,inits = NmodInits) 

combinedchains <- mcmc.list(posterior$samples$chain1, 
                            posterior$samples$chain2)

png("combinedchains.png", width = 1200, height = 800)
plot(combinedchains)
dev.off()

autocorr.plot(posterior$samples$chain1)

autocorr.plot(posterior$samples$chain2)
```

```{r}
gelman.diag(combinedchains)
```

```{r}
gelman.plot(combinedchains)
```

```{r}
posterior$summary$all.chains
```

```{r}
eff_s_s_mu1 <- effectiveSize(posterior$samples$chain1[ , "mu1"])
cat("The effective sample size for mu for chain1 is",eff_s_s_mu1)
```

```{r}
eff_s_s_diff <- effectiveSize(posterior$samples$chain1[ , "diff"])
cat("The effective sample size for the difference in means for chain1 is",eff_s_s_diff)
```

```{r}
eff_s_s_tau <- effectiveSize(posterior$samples$chain1[ , "tau"])
cat("The effective sample size for tau for chain1 is",eff_s_s_tau)
```

```{r}
batchSE(posterior$samples$chain1,batchSize=100) # For MC errors (only one chain is used.)
```