---
title: "Tutorial5"
format: html
editor: 
  markdown: 
    wrap: 80
---

# Packages and Data

```{r, message= F}
library(tidyverse)
library(nimble)
library(coda)
library(igraph)
library(R6)
library(knitr)
```

Consider birth weight data ğ‘¤ = (ğ‘¤1, ..., ğ‘¤200), in Kg, from 200 births at
Hospital H. The data are given in the R workspace file Birth_weights_Tutorial_3
in the Tutorials folder in Moodle. The data file also contains binary
information on the Index of Multiple Deprivation (IMD) of the mother. 0 denotes
high deprivation (index from 1 to 5), while 1 denotes low deprivation (index
from 6 to 10). Note the counter-intuitive interpretation of the IMD values on
the scale from 1 to 10, and subsequently on the binary scale.

```{r, message=FALSE}
load("Birth_weights_Tutorial_5.RData")

d <- as.data.frame(cbind(age, birth_weights, IMD_binary, pm2_5))

d <- d |>
  rename(
    bw = birth_weights,
    imd = IMD_binary) |>
  mutate(
    cage = age - mean(age),
    cpm2_5 = pm2_5 - mean(pm2_5),
    imd_plot = ifelse(imd == 1, "Low", "High"))

# remove(age, birth_weights, cage, cpm2_5, imd, IMD_binary)
```

Fit a linear model with the birth weights ğ‘¤ğ‘– , ğ‘– = 1, ..., 200 as the outcome
that follows a Normal distribution so that (given ğœ‡ğ‘– and ğœ), (ğ‘¤ğ‘– \|ğœ‡ğ‘– , ğœ âˆ¼ ğ‘(ğœ‡ğ‘–
, ğœ2 )), with the weights being independent. Center the continuous covariates
age and pm2_5, creating two new covariates, say, **cage** and **cpm2_5**
Overall, the covariates should be the binary IMD variable, as well as the
centered **cage** and **cpm2_5**.

Parametrise the mean ğœ‡ğ‘– for birth weight ğ‘¤ğ‘– so that,

ğœ‡ğ‘– = ğ›½0 + ğ›½1 âˆ— ğ¼ğ‘€ğ·ğ‘– + ğ›½2 âˆ— ğ‘ğ‘ğ‘”ğ‘’ + ğ›½3 âˆ— ğ‘ğ‘ğ‘š2_5.

Place a ğ¿ğ‘œğ‘” âˆ’ ğ‘ğ‘œğ‘Ÿğ‘šğ‘ğ‘™(1, 0.72 ) prior on ğ›½0 . Place suitably non-informative
priors on ğ›½1 , ğ›½2 and ğ›½3 . For the precision ğœ = 1/ğœ2 , we will assume a Gamma
prior distribution ğº(0.1, 0.1).

Use Nimble to derive the posterior distribution ğœ‹(ğ›½0 , ğ›½1 , ğ›½2 , ğ›½3 , ğœ2 \|ğ‘¤).
Run 2 chains. Consider a burn-in of 100000 iterations, with 400000 further
iterations for each chain. Thin by 10.

Obtain plots of the posterior distributions and diagnostic plots. Check the MCMC
convergence, not forgetting to check the chains length in relation to the
autocorrelations and MC error.

Answer the following questions, to an accuracy of 3 decimal places where
appropriate. (Note that you should also be checking model assumptions and model
fit before reporting inferences, but this will be one of the topics in the next
tutorial sheet.)

```{r}
#covariates and response

y = d$bw; imd = d$imd; cage = d$cage; cpm2_5 = d$cpm2_5

# Specify the statistical model

n = nrow(d)

mod <- nimbleCode({

# Specify the likelihood:
for(i in 1:n){
  
  mu[i] <- beta0+beta1*imd[i]+beta2*cage[i]+beta3*cpm2_5[i]
  
  y[i]~dnorm(mu[i],tau)
  
}

# Prior for beta

beta0~dnorm(mu0,tau0)
beta1~dnorm(mu0,tau0)
beta2~dnorm(mu0,tau0)
beta3~dnorm(mu0,tau0)

tau0 <- 1/sigma02

# Prior for the precision
tau~dgamma(a, b)

# Compute the variance
sigma2 <- 1/tau
})
```

```{r}
# hyperparameters for the betas and tau
# would find n in R environment. Including here to suppress a warning
Consts <- list(mu0=0, sigma02=1000, a=0.1, b=0.1, n=200)

# list with data
Data <- list(y=y,imd=imd,cage=cage,cpm2_5=cpm2_5)

# Initial values
Inits <- list(beta0 = 1, beta1 = 0, beta2 = 0, beta3 = 0, tau = 1) 
# Nimble will generate the rest 

# Build and Compile the model
# to build the model
mod_re <- nimbleModel(code = mod, name = "mod", constants = Consts,
                    data = Data, inits<-Inits)

# To compile the model
Cmod_re <- compileNimble(mod_re)
```

```{r}
mod_Conf <- 
  configureMCMC(mod_re,
                monitors = c("beta0", "beta1", "beta2", "beta3", "sigma2", "tau"),
                enableWAIC = TRUE, print = TRUE) 
```

```{r}
# build the MCMC algorithm
mod_reMCMC <- buildMCMC(mod_Conf)
# compile the MCMC chain 
Cmod_reMCMC <- compileNimble(mod_reMCMC, project = mod_re)
```

```{r}
####################################################################################
####### POSTERIOR SAMPLES IN CODA FORMAT TO GET MORE EASILY PLOTS AND DIAGNOSTICS  #
####################################################################################
set.seed(10)

mod_reInits <- list(list(beta0 = 1, beta1 = 0, beta2 = 0, beta3 = 0, tau = 1), 
                     list(beta0 = 2, beta1 = 0, beta2 = 0, beta3 = 0, tau = 0.5))

posterior <- runMCMC(Cmod_reMCMC, niter = 500000, thin=10, nburnin=100000, 
                     summary = TRUE, WAIC = TRUE, samples = TRUE, nchains=2, 
                     samplesAsCodaMCMC=TRUE, inits = mod_reInits) 

combinedchains <- mcmc.list(posterior$samples$chain1, posterior$samples$chain2)
```

```{r}
summary(combinedchains)
```

\

```{r}
plot(combinedchains) # too many plots sometimes
```

```{r}
#plot(combinedchains[,c('beta1','beta2','sigma2')]) 
#plot(combinedchains[,'beta[1]']) # when defining a vector of parameters

autocorr.plot(posterior$samples$chain1)
autocorr.plot(posterior$samples$chain1[,"tau"])

autocorr.plot(posterior$samples$chain2)
```

```{r}
gelman.diag(combinedchains)
gelman.plot(combinedchains)
```

```{r}
ESS <- t(effectiveSize(combinedchains))
kable(ESS, digits = 3)
# cat("The Effective Sample Size is ", ESS)
```

a\) Posterior inferences can be trusted, considering convergence and mixing. Is
this statement true or false? \[1 mark\]

b)  Considering births at hospital H in general, state the estimated mean birth
    weight for parents classified within then high-deprivation band, of average
    age, with an exposure to PM2.5 that is also average. \[1 mark\]

c)  Provide the left limit of the 95% credible interval for the estimate in (b).
    \[1 mark\]

d)  Provide the right limit of the 95% credible interval for the estimate in
    (b). \[1 mark\]

e)  State the predicted weight for a specific birth from a mother that is
    classified within the low-deprivation band, is 29 years old, and their
    exposure to PM2.5 is 16 ğœ‡ğ‘”/ğ‘š3. \[1mark\]

f)  Provide the left limit of the 95% credible interval for the prediction in
    (e). \[1 mark\]

g)  Provide the right limit of the 95% credible interval for the prediction in
    (e). \[1 mark\]

h)  Describe to someone with no statistical knowledge the effect of the model
    covariates, considering the interpretation of the estimates for ğ›½0, ğ›½1, ğ›½2,
    ğ›½3. \[3 marks\]

i)  Comment on how important each one of the covariates seems to be for the
    modelling of birth weights. \[2 marks\]
