---
title: "Week1"
format: html
---

```{r}
# install.packages("nimble")
# install.packages("igraph")
# install.packages("coda")
# install.packages("R6")
# writeLines('PATH="${RTOOLS40_HOME}\\usr\\bin;${PATH}"', con = "~/.Renviron")
```

```{r}
library(nimble)
library(igraph)
library(coda)
library(R6)
```

Simple code to use Nimble

At this stage we will not use Nimble to analyse data or fit a statistical model. We will just
use it to sample from well-known distributions.


Sample from a Normal distribution

We will pretend that the ‘rnorm()’ R function is not available. We will use Nimble to sample
1000 realisations/deviates from a 𝑁 (0, 32) distribution. (This is an overkill really! Nimble is
supposed to be used for more complex tasks.) Use the following code, to also plot a histogram
of the sample. Do not worry about knowing what each line does exactly. We will
discuss this in detail in Chapter 2.

```{r}
# Model
sim1code <- nimbleCode({
y ~ dnorm(0, sd = 3)
})
# Build model
sim1Model <- nimbleModel(sim1code)
# To compile the model
Csim1Model <- compileNimble(sim1Model)
# set up monitoring for y
sim1ModelConf <- configureMCMC(sim1Model, monitors = c('y'), print = TRUE)
# build the algorithm
sim1ModelMCMC <- buildMCMC(sim1ModelConf)
# compile the sampler
Csim1ModelMCMC <- compileNimble(sim1ModelMCMC, project = sim1Model)
set.seed(1)
sample1 <- runMCMC(Csim1ModelMCMC, niter = 1000, thin=1, nburnin=0,
summary = TRUE, samples = TRUE, nchains=1)
hist(sample1$samples, main="1000 realisations from N(0,9)",
ylab="Number of Samples")
```

Sample from a Binomial distribution

We will pretend that the ‘rbinom()’ R function is not available, and use Nimble (again overkill!)
to sample 1000 realisations/deviates from a 𝐵𝑖𝑛(10, 0.8) distribution. Use the following code,
to also obtain a bar-chart of the sample:

```{r}
# Model
sim2code <- nimbleCode({
x ~ dbin(0.8, 10)
})
# Build model
sim2Model <- nimbleModel(sim2code)
# To compile the model
Csim2Model <- compileNimble(sim2Model)
# set up monitoring for x
sim2ModelConf <- configureMCMC(sim2Model, monitors = c('x'), print = TRUE)
# build the algorithm
sim2ModelMCMC <- buildMCMC(sim2ModelConf)
# compile the sampler
Csim2ModelMCMC <- compileNimble(sim2ModelMCMC, project = sim2Model)
set.seed(1)
sample2 <- runMCMC(Csim2ModelMCMC, niter = 1000, thin=1, nburnin=0,
summary = TRUE, samples = TRUE, nchains=1)
# Obtain plot
plot_sample <- as.vector(table(sample2$samples))
plot_min <- min(as.integer(names(table(sample2$samples))))
plot_max <- max(as.integer(names(table(sample2$samples))))
plot(plot_min:plot_max,plot_sample, type="h",
main="1000 realisations from Bin(10,0.8)",
xlab="Sampled value", ylab="Number of Samples")
```

Exercise


Plotting the samples, use Nimble to sample 1000 realisations from a
1. 𝑃 𝑜𝑖𝑠𝑠𝑜𝑛(3) distribution,
2. 𝐺𝑎𝑚𝑚𝑎(2, 3) distribution.



Hint: In the code provided above, the only important change to make is to the distribution
specified within the specification of the model. Use the Nimble manual (available in the Nimble
website mentioned above) to see how different distributions can be specified.

```{r}
# Model
sim2code <- nimbleCode({
x ~ dpois(3)
})
# Build model
sim2Model <- nimbleModel(sim2code)
# To compile the model
Csim2Model <- compileNimble(sim2Model)
# set up monitoring for x
sim2ModelConf <- configureMCMC(sim2Model, monitors = c('x'), print = TRUE)
# build the algorithm
sim2ModelMCMC <- buildMCMC(sim2ModelConf)
# compile the sampler
Csim2ModelMCMC <- compileNimble(sim2ModelMCMC, project = sim2Model)
set.seed(1)
sample2 <- runMCMC(Csim2ModelMCMC, niter = 1000, thin=1, nburnin=0,
summary = TRUE, samples = TRUE, nchains=1)
#Obtain plot
plot_sample <- as.vector(table(sample2$samples))
# plot_min <- min(as.integer(names(table(sample2$samples))))
# plot_max <- max(as.integer(names(table(sample2$samples))))
plot(plot_sample, type="h",
main="1000 realisations from Pois(10,0.8)",
xlab="Sampled value", ylab="Number of Samples")

plot_sample
```

```{r}
# Model
sim2code <- nimbleCode({
x ~ dgamma(2, 3)
})
# Build model
sim2Model <- nimbleModel(sim2code)
# To compile the model
Csim2Model <- compileNimble(sim2Model)
# set up monitoring for x
sim2ModelConf <- configureMCMC(sim2Model, monitors = c('x'), print = TRUE)
# build the algorithm
sim2ModelMCMC <- buildMCMC(sim2ModelConf)
# compile the sampler
Csim2ModelMCMC <- compileNimble(sim2ModelMCMC, project = sim2Model)
set.seed(1)
sample2 <- runMCMC(Csim2ModelMCMC, niter = 1000, thin=1, nburnin=10,
summary = TRUE, samples = TRUE, nchains=1)
#Obtain plot
# plot_sample <- as.vector(table(sample2$samples))
# # plot_min <- min(as.integer(names(table(sample2$samples))))
# # plot_max <- max(as.integer(names(table(sample2$samples))))
# plot(plot_sample, type="h",
# main="1000 realisations from Gamma(2, 3)",
# xlab="Sampled value", ylab="Number of Samples")
# 
# plot_sample

hist(sample2$samples)
```

